---
title: "Lipidomic Analysis - Relative abundance Analysis"
author: "Aiub Mohamed Barara"
date: "2025-07-01"
output:
  html_document: default
  pdf_document: default
---

```{r}
knitr::opts_chunk$set(dev = "png", dpi = 600, fig.width = 6, fig.height = 4)
```

```{r setup, include=FALSE}
# Required packages
library(lipidr)
library(ggplot2)
library(pheatmap)
library(corrplot)
library(readxl)
library(dplyr)
library(FactoMineR)
library(factoextra)
library(tidyverse)
library(fmsb)
```

# Data Import & Preprocessing

## Data Wrapping into Lipidomics Experiment Object

```{r}
# We load the relative abundance lipidomics data 

archivo_excel <- "Lipidomics_v2.xlsx"

percent_total <- read_excel(archivo_excel, sheet = "%_total")
```

```{r}

# We convert the imported data into a lipidomics experiment object
# This facilitates standardized downstream analysis 
d <- as_lipidomics_experiment(percent_total)
names(assays(d)) <- "relative_abundance_percent" # We assign a name to the assay
metadata(d)$units <- "relative_percent" # We specify the unit of measurement
metadata(d)$description <- "Percentage values relative to total lipid abundance" # We annotate a descriptive metadata entry
```

## Sample Annotation

```{r}
# We load the sample annotation file and incorporate it into the experiment object

annot <- read_excel("data_lipid_clin.xlsx")
d <- add_sample_annotation(d, annot)

# We convert sample group information into a factor for downstream grouping

colData(d)$SampleType <- as.factor(colData(d)$SampleType)
annotation_col <- data.frame(Group = colData(d)$SampleType) # We create a simplified annotation dataframe
```

```{r}
# We view the full set of 22 column names in the rowData object
colnames(rowData(d))
```

# Descriptive analysis

## Quality Control

```{r}
# We assess whether relative abundances per sample approximately sum to 100%, which is expected given percentage normalization

colSums(assay(d, "relative_abundance_percent"))
```

```{r}
# We convert the matrix of lipid measurements into a data frame
# and add lipid names as a column to enable reshaping
df <- as.data.frame(assay(d, "relative_abundance_percent"))
df$Lipid <- rownames(df)

# We convert the data to long format to facilitate ggplot visualization
df_long <- tidyr::gather(df, key = "Sample", value = "Relative_Abundance", -Lipid)

# We extract the group information from sample names by removing trailing numbers
df_long$Group <- gsub("-\\d+", "", df_long$Sample)
df_long$Group <- factor(df_long$Group, levels = c("WT", "ntrc", "OE"))
```

```{r}
# We generate a global boxplot across all lipid species to visualize overall distribution by group
ggplot(df_long, aes(x = Group, y = Relative_Abundance, fill = Group)) + 
  geom_boxplot(outlier.size = 0.5, alpha = 0.8) + 
  scale_y_log10() +
  labs(title = "Distribución de abundancia relativa de lípidos por condición",
    x = "Condición",
    y = "Abundancia relativa (%)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

```{r}
# We plot lipid class distribution across all samples
plot_lipidclass(d, type = "boxplot", measure = "relative_abundance_percent",  log = T) +
  ggtitle("Distribución por clase lipídica (relative abundance)")
```

```{r}
## Visualization: Per-Condition Boxplots
# We define the experimental groups
condiciones <- c("WT", "ntrc", "OE")

# We generate a lipid class boxplot per condition
for (cond in condiciones) {
  # Filtrar muestras de la condición actual
  d_filtrado <- d[, grepl(paste0("^", cond), colnames(d))]
  
  # Gráfico para la condición
  p <- plot_lipidclass(d_filtrado, type = "boxplot", 
    measure = "relative_abundance_percent",  
    log = TRUE) +
    ggtitle(paste("Lipid Classes on", cond, "relative_abundance_percent")) +
    theme(plot.title = element_text(hjust = 0.5))
  
  print(p)
}
```

## Principal Component Analysis

```{r}
# We perform principal component analysis to evaluate sample variation based on lipid profiles
mva_res <- mva(d, measure = "relative_abundance_percent", method = "PCA")
plot_mva(mva_res, color_by = "SampleType", components = c(1, 2)) +
  ggtitle("PCA de perfiles moleculares (abundancias relativas)")
```

## Heatmap of Lipid Species

```{r}
# We transform the abundance matrix and apply log2 transformation for better visualization
mat <- percent_total %>%
  column_to_rownames("Compound Name") %>%
  as.matrix()

# We apply a +1 transformation to avoid log(0)
mat_log <- log2(mat + 1)
```

```{r}
# We visualize lipid abundance patterns using a clustered heatmap
pheatmap(mat_log, 
         cluster_rows = TRUE, 
         cluster_cols = TRUE, 
         scale = "row",           
         clustering_distance_rows = "correlation",
         clustering_method = "complete",
         main = "Heatmap de abundancia relativa (log2 transformada)",
         fontsize_col = 12,      
         fontsize_row = 6)      
```

# Differential relative abundance analysis

## Differential Analysis: ntrc vs WT

```{r}
# We perform differential analysis using limma, contrasting ntrc against WT

de_results_ntrc_WT <- de_analysis(d,ntrc - WT,measure = "relative_abundance_percent")
significant_molecules(de_results_ntrc_WT) # We retrieve significantly different molecules
plot_results_volcano(de_results_ntrc_WT, show.labels = FALSE) # We generate a volcano plot
```

## Differential Analysis: OE vs WT

```{r, fig.height=6}
de_results_OE_WT <- de_analysis(d,OE - WT,measure = "relative_abundance_percent")
significant_molecules(de_results_OE_WT)
plot_results_volcano(de_results_OE_WT, show.labels = FALSE)
```

## Differential Analysis: ntrc vs OE

```{r}
de_results_ntrc_OE <- de_analysis(d,ntrc - OE,measure = "relative_abundance_percent")
significant_molecules(de_results_ntrc_OE)
plot_results_volcano(de_results_ntrc_OE, show.labels = FALSE)
```

# Lipid Class Analysis (relative abundance)

```{r}
# We load lipid class percentage values
Percent_Class <- read_excel(archivo_excel, sheet = "%_Class")

head(Percent_Class)
```

```{r}
# We reshape the class-level data to long format for plotting
# We annotate group membership based on sample naming conventions
df_long <- Percent_Class %>%
  pivot_longer(-Class, names_to = "Sample", values_to = "Relative_Abundance") %>%
  mutate(Group = str_remove(Sample, "-\\d+")) %>%
  mutate(Group = factor(Group, levels = c("WT", "ntrc", "OE")))
```

```{r}
# We generate a boxplot distribution by lipid class 
ggplot(df_long, aes(x = Group, y = Relative_Abundance, fill = Group)) +
  geom_boxplot(alpha = 0.8, outlier.size = 0.5) +
  facet_wrap(~ Class, scales = "free_y") +
  labs(
  title = "Distribution of Relative Abundance by Lipid Class",
  x = "Genotype",
  y = "Relative Abundance (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

```{r}
# We reshape to wide format to calculate correlation between lipid classes
df_wide <- Percent_Class %>%
  pivot_longer(-Class, names_to = "Sample", values_to = "Value") %>%
  pivot_wider(names_from = Class, values_from = Value)

mat_class <- df_wide %>%
  select(-Sample) %>%
  as.matrix()

# We compute Pearson correlation across lipid classes
cor_mat <- cor(mat_class, method = "pearson")
```

```{r}
# We visualize the correlation matrix using corrplot
corrplot(cor_mat, method = "color", 
         type = "upper", 
         addCoef.col = "black", 
         tl.col = "black", 
         number.cex = 0.7, 
         title = "Correlación entre clases lipídicas")
```

```{r}
# We prepare data for radar chart analysis to visualize mean values by group
df_long <- Percent_Class %>%
  pivot_longer(-Class, names_to = "Sample", values_to = "Value") %>%
  mutate(Group = gsub("-\\d+", "", Sample))

# We compute group-level means across lipid classes
df_means <- df_long %>%
  group_by(Group, Class) %>%
  summarise(MeanValue = mean(Value, na.rm = TRUE)) %>%
  pivot_wider(names_from = Class, values_from = MeanValue)

# We convert to matrix format required by fmsb
radar_data <- as.data.frame(df_means[,-1])
rownames(radar_data) <- df_means$Group

# We add rows with max and min values for radar chart scaling
max_vals <- apply(radar_data, 2, max)
min_vals <- rep(0, ncol(radar_data))
radar_data <- rbind(max_vals, min_vals, radar_data)
```

```{r}
# We define color scheme for radar chart
colors_border <- c("red", "blue", "green")
colors_in <- scales::alpha(colors_border, 0.3)
```

```{r}
# Finally, we draw the radar chart
radarchart(radar_data, axistype=0,
           pcol=colors_border, pfcol=colors_in, plwd=2, plty=1,
           cglcol="grey", cglty=1, cglwd=0.8,vlcex=0.8)

legend(x=1, y=1, legend=rownames(radar_data)[-c(1,2)], bty="n", pch=20 , col=colors_border, text.col = "black", cex=0.8, pt.cex=1.5)
```
